TacticToe: Learning the Game of Tactics

### 2018-01-29 tactictoe_jv.tex
adapting orthogonalization and abstraction

\section{Orthogonalization}\label{sec:ortho}
Different tactics may transform a single goal in the same way. Exploring such 
equivalent paths
is undesirable, as it leads to inefficiency in automated proof search.
To solve this problem, we do not directly assign a goal to the associated 
tactic, but organize a competition on the closest feature vectors. The number 
of selected feature vectors is a parameter that is set for each experiment.
The winner is the tactic appearing in the most feature vectors provided that it 
has the 
same effect as the original tactic. We associate this tactic with the features 
of the targeted goal instead of the original in our feature database. As a 
result, already successful tactics are preferred, and new tactics are 
considered only if they provide a different contribution. We also advantage 
tactics that have a large coverage. In the following, we give formal definition 
of the concepts of subsumption and coverage that are required for expressing 
the orthogonalization algorithm.

@@

\begin{definition} (Goal subsumption)\\ 
A goal subsumption $S$ is a partial relation on a set of goals.\\
Given two goals $g_1$ and $g_2$, we note $g_1 \ge g_2$ when  $(g_1,g_2) \in 
S$.\\
Assuming we have a way to test if a goal is a consequence of another, a 
possibly useful subsumption definition would be: 
\[g_1 \ge g_2  \Leftrightarrow_{def} g_1 \mbox{ is a consequence of } g_2\]
In our experiments, we take a minimal subsumption defined by:
\[g_1 \ge g_2  \Leftrightarrow_{def} g_1 \mbox{ is }\alpha\mbox{-equivalent 
to } g_2\]
\end{definition}

We can naturally extend a goal subsumption to a set of goals.
\begin{definition} (Goal list subsumption)\\
A goal set subsumption is an partial relation on a set of goals.
Given two list of goals $l_1$ and $l_2$, we can define it from goal 
subsumption $\ge$ by:
\[l_1 \ge l_2  \Leftrightarrow_{def} \forall g_2 \in l_2\ \exists g_1 \in l_1\ 
g_1 \ge g_2\]
\end{definition}

We now build a subsumption on the tactic level.
\begin{definition} (Tactic subsumption)\\  
Given two non-failing tactics $t_1$ and $t_2$ on $g$, a tactic $t_1$ subsumes a 
tactic $t_2$ on a goal $g$, noted $\ge_g$, when:

@@

If one of the tactic is failing then the couple $(t_1,t_2)$ is not an element 
of the 
relation.
Given the minimal subsumption on goals, a tactic $t_1$ subsumes another $t_2$ 
on a goal 
$g$ if and only if the resulting set of goals produced by $t_1$ is a subset 
of the output of $t_2$ on $g$. 
\end{definition}

We can now express the orthogonalization algorithm given in a single formula 
with parameters: $P_n$ for the tactic predictor where $n$ for the number of 
predicted
tactics, $S$ for the tactic subsumption relation and $\textit{\mbox{Coverage}}$ 
for the coverage of 
a tactic. The orthogonalization process $\mathbb{O}$ is given as input a 
recorded feature vector $(t,g)$ where $t$ is the tactic originally applied to 
the goal $g$. The winning tactic of the competition is given by the formula:

@@

After orthogonalization, because the orthogonalized tactic is supposedly 
better, we record the feature vector $(\mathbb{O}(t,g),g)$ instead of 
$(t,g)$. To sum up, the winning tactic is the one with the most coverage and 
that is able to replicate the effect of the original tactic.

\section{Abstraction}\label{sec:synthesis}\todo{MN + RK: review + fix typos}
One of the great weakness of the previous version of \tactictoe is that
it could not create its own tactics. Indeed, sometimes no previous tactic is 
adapted for a certain goal and creating a new tactic is necessary.
%Experiments in section, show how different degrees of tactic synthesis could 
%improve the success rate of \tactictoe.
We present a way to create tactics with different arguments
by abstracting them and re-instantiating them using a predictor for that kind 
of argument. In the spirit of the orthogonalization method, we will try to 
create tactics that are more general but have the same effect as the 
ungeneralized one.

@@

\paragraph{Instantiation of an Abstracted Tactic}
An abstracted tactic is not immediately applicable to a goal, since it contains
unspecified arguments. To apply an abstracted tactic $A$, we first need to 
instantiate that should placeholders $h$ inside $A$. Because it is difficult to 
build a general predictor that effective on each type of argument, we manually 
design different argument predictors for each type. Those predictors are 
usually given a goal has input and returns the best chosen argument for this 
goal(see Section~\ref{sec:arg_pred}). All in all, an abstracted tactics has 
dynamic arguments, it decides how to instantiate them depending on the given 
goal.

\paragraph{Selection of Abstracted Tactics}
As abstracted tactics are not used in human proofs, we need to find a way to 
predict them so that they can contribute to the \tactictoe proof search.
A straightforward idea is to duplicate each tactic during the proof search by 
creating an abstracted version of the tactic. However, this is computationally 
expensive when the tactic does not instantiate back to its original form, we 
risk doing unnecessary work as both version may have a similar effect.
Therefore, we need to decide beforehand which of the two is better.
In fact, we can re-use the orthogonalization module to do this task for us.
During the orthogonalization competition on goal $g$, before trying a tactic 
$T$, we now also try to apply its abstracted version $A$. If $A$ succeeds, it 
becomes a a label for $g$ is including in the database of tactic features and 
thus can be predicted during a proof search.

\begin{remark}
Forcing $A$ into the competition is necessary because it has never been seen 
before so it has no coverage. But after the coverage $c_A$ of $A$ in the 
database exceeds the coverage of $T$, $A$ will be selected before $T$ without 
the previously required forcing.
\end{remark}

We made argument predictors for one type of arguments: list 
of theorems. We choose it is the most common type and the structure of theorems 
are complex so they contain enough information to be predicted accurately. 
We plan to extend our algorithm to term which is another frequent argument 
types but term synthesis is so far too weak to positively 
influence the proof search.

>>>>

\section{Orthogonalization}\label{sec:ortho}
Different tactics may transform a single goal in the same way. Exploring such 
equivalent paths
is undesirable, as it leads to inefficiency in automated proof search.
To solve this problem, we do not directly assign an input goal $g$ to its 
associated tactic $t$, but organize a competition between the $n$ closest 
tactic-goal pairs, noted $P_n(g)$. 
The winner is the tactic that subsumes the original tactic on $g$ and that
appears in the largest number of tactic-goal pairs in the database.
The winning tactic $w$ is then associated with $g$, producing the pair $(w,g)$ 
that replaces the original pair $(t,g)$ in the database. As a 
result, already successful tactics with a large coverage are preferred, and new 
tactics are 
considered only if they provide a different contribution. In the following, we 
give a formal definition of the concepts of subsumption and coverage that are 
required for expressing the orthogonalization algorithm.

@@

\begin{definition} (Goal subsumption)\\ 
A goal subsumption $S$ is a partial relation on a set of goals.\\
Given two goals $g_1$ and $g_2$, we note $g_1 \ge g_2$ when  $(g_1,g_2) \in 
S$.\\
Assuming we have a way to test if a goal is a consequence of another, a 
possibly useful subsumption definition would be: 
\[g_1 \ge g_2  \Leftrightarrow_{def} g_1 \mbox{ is a consequence of } g_2\]
By default, we choose a minimal subsumption defined by:
\[g_1 \ge g_2  \Leftrightarrow_{def} g_1 \mbox{ is }\alpha\mbox{-equivalent 
to } g_2\]
\end{definition}

We can naturally extend a goal subsumption to lists of goals.
\begin{definition} (Goal list subsumption)\\
A goal list subsumption is an partial relation on lists of goals.
Given two list of goals $l_1$ and $l_2$, we can define it from goal 
subsumption $\ge$ by:
\[l_1 \ge l_2  \Leftrightarrow_{def} \forall g_2 \in l_2\ \exists g_1 \in l_1\ 
g_1 \ge g_2\]
\end{definition}

This allows us to define subsumption for tactics.
\begin{definition}(Tactic subsumption)\\  
Given two non-failing tactics $t_1$ and $t_2$ on $g$, a tactic $t_1$ subsumes a 
tactic $t_2$ on a goal $g$, noted $\ge_g$, when:

@@

If one of the tactic is failing then $t_1$ and $t_2$ are not comparable through 
this relation.
Given the default subsumption on goals, $t_1$ subsumes $t_2$ on $g$ if and only 
if $t_1(g)$ is a subset (modulo $\alpha$-equivalence) of $t_2(g)$. 
\end{definition}

In the end, the winning tactic of the orthogonalization competition can be 
expressed by the formula:

@@

As a consequence $(\mathbb{O}(t,g),g)$ replaces $(t,g)$ in the database of 
tactics.

\section{Abstraction}\label{sec:synthesis}
One of the great weakness of the previous version of \tactictoe is that
it could not create its own tactics. Indeed, sometimes no previous tactic is 
adapted for a certain goal and creating a new tactic is necessary.
%Experiments in section, show how different degrees of tactic synthesis could 
%improve the success rate of \tactictoe.
We present a way to create tactics with different arguments
by abstracting them and re-instantiating them using a predictor for that kind 
of argument. In the spirit of the orthogonalization method, we will try to 
create tactics that are more general but have the same effect as the 
ungeneralized one.

@@

\paragraph{Instantiation of an Abstracted Tactic}
An abstracted tactic is not immediately applicable to a goal, since it contains
unspecified arguments. To apply an abstracted tactic $A$, we first need to 
instantiate the placeholders $h$ inside $A$. Because it is difficult to 
build a general predictor that effective on each type of argument, we manually 
design different argument predictors for each type. Those predictors are 
usually given a goal has input and returns the best chosen argument for this 
goal. All in all, an abstracted tactics has 
dynamical arguments, it instantiate them differently for different goals.

We made successful argument predictors for one type of arguments: list 
of theorems. It is a very common type and the structure of theorems 
are complex, thus they contain enough information to be predicted accurately. 
We plan to extend our algorithm to term which is another frequent argument 
types but term synthesis is so far too weak to positively 
influence the proof search.

\paragraph{Selection of Abstracted Tactics}
As abstracted tactics do not appear in human proof, we need to find a way to 
predict them so that they can contribute to the \tactictoe proof search.
A straightforward idea is to duplicate each tactic during the proof search by 
creating an abstracted version of the tactic. However, this is computationally 
expensive when the tactic does not instantiate back to its original form, we 
risk doing unnecessary work as both version may have a similar effect.
Therefore, we need to decide beforehand which of the two is better.
In fact, we can re-use the orthogonalization module to do this task for us.
We add $A$ to the competition, initially giving it the coverage of $T$. If $A$ 
wins, it becomes a a label for $g$ is including in the database of tactic 
features and thus can be predicted during a proof search.
After many competitions, the coverage of $A$ may exceed the coverage of $T$. At 
this point, the coverage of $A$ is estimated on its own and not inherited from 
$T$ anymore.
